Learning a functional distance between organizations with unsupervised embeddings of scientific mobility

### 2020-07-24 paper/current/main.tex
update 2nd para
edit final main para

Just as the embedding space makes it possible to ``zoom in'' on subsets of organizations, it is also possible to ``zoom out'' by aggregating organizational vectors. 
For example, we can examine the high-level structure by estimating the aggregate embeddings of countries. 
Namely, we define the representative vector of each country as the average of their organizational vectors.
We then perform hierarchical clustering of nations that have at least 25 organizations represented in the embedding space, excluding the United States which is a dominant hub well connected to most other countries (see Fig. \ref{fig:clustering}a). 
The six identified clusters roughly correspond to countries in East-Asia (orange), Scandinavia (dark purple), the British Commonwealth (light purple), and Central and Eastern Europe (light blue), and then two remaining diverse clusters contain a mix of European, Latin American, and Mediterranean countries.
These clusters demonstrate the importance of not only geography but also linguistic ties between countries.

@@

Prestige hierarchy is known to underpin the dynamics of scientific mobility~\autocite{deville2014career, clauset2015hierarchy}, but can this hierarchy be encoded in the embedding space, to which no explicit information on prestige is given?
We test this by exploiting the geometric properties of the embedding space with SemAxis~\autocite{an2018semaxis}, an approach originally designed to characterize the semantics of words in word embeddings.
Here, we use SemAxis to operationalize the abstract notion of academic prestige, defining each pole of the axis using high- and low-ranked universities. 
We define prestige using the Times Ranking of World Universities (and using research impact from the Leiden Rankings~\autocite{waltman2012leidenrankings} in Supplementary Materials), including all ranked universities in the United States;
the high-ranked pole is defined as the mean vector of the top five U.S. universities according to the rankings, whereas the low-ranked pole is defined using the five bottom-ranked and geographically matched (by U.S. census region) universities.
A geographic axis is also added to aid the visualization and reinforce the importance of geography, defined as the mean vector of all universities in California and then Massachusetts.

@@

The embedding space captures the prestige hierarchy and can be used to reconstruct academic rankings. 
We derive a ranking for universities using the SemAxis prestige axis in Fig.~\ref{fig:semaxis}a and compare them to traditional university rankings.
The SemAxis ranking is defined from the position of organizations along the prestige axis, with the top-ranked being the furthest towards the high-ranked end of the axis.
We find that the embedding-based ranking is strongly correlated with the Times ranking (Spearman's $\rho = 0.73$, Fig.~\ref{fig:semaxis}e).
The strongest outliers that were ranked more highly in the Times Ranking than the SemAxis ranking tend to be large state universities such as Arizona State University and  University of Florida;
Those ranked higher in the SemAxis ranking tend to be relatively-smaller universities such as the University of San Francisco and University of Maryland Baltimore County that are close to other prestigious universities.

The correlation between SemAxis and the Times ranking is robust with the number of organizations used to define the axes (Fig. SI~\ref{fig:supp:semaxis_compare}), such that even using only the single top-ranked and bottom-ranked universities are significantly correlated with the Times Ranking (Spearman's $\rho = 0.46$, Fig.~\ref{fig:supp:semaxis_compare}a).
For reference, the SemAxis-derived ranking achieves correlation with the Times Ranking comparable to that between the Times ranking and the Leiden Ranking~\autocite{waltman2012leidenrankings}, a well-known and bibliometrically-informed university ranking (Spearman's $\rho = 0.87$, Fig.~\ref{fig:semaxis}f).
This suggests that the organization embedding captures the latent prestige hierarchy inherent to academic mobility, even without providing any explicit information on prestige, with a high degree of resolution and reliability~\autocite{deville2014career, clauset2015hierarchy}.

What is it about the organization embedding that captures prestige so effectively?
We find that the size (L2 norm) of the organization embedding vectors provides insights into the characteristics of organizations. 
Up to a point, the size of U.S. organization's vectors tends to increase proportional to its number of researchers (both mobile and non-mobile) who published with that organization;
these organizations are primarily teaching-focused institutions, agencies, and hospitals that either are not ranked or have a low ranking. 
However, at around 1,000 researchers, the size of the vector decreases as the number of researchers increases;
organizations here are primarily research-intensive and prestigious universities with higher rankings, research outputs, R\&D funding, and doctoral students (Fig.~\ref{fig:supp:length_vs_metaInfo}).
A similar pattern has been observed in applications of neural embedding to natural language, wherein the size of word vectors were found to represent word ``significance''~\autocite{schakel2015measuring}.
If a word appears frequently in a specific context, the norm of the word vector is likely to be large, because a large norm is associated with how predictably the word occurs given a context.
However, if the word in question is universal, appearing frequently in many different contexts, it cannot have a large norm as it lacks a strong association with a particular context. 
Likewise, an organization such as Harvard with a small norm appears in many contexts, alongside many different organizations in affiliation trajectories. 
The concavity emerges in part from the relationship between the size of the vector and the expected connectedness of the organization given its size (Fig.~\ref{fig:length}a, inset, $R^{2} = 0.17$).
Large, prestigious, and well-funded research universities such as Princeton and Harvard have smaller vector sizes because they appear in many different contexts compared to more teaching-focused organizations such as NY Medical College, and University of Michigan at Flint. 
Some universities, such as the University of Alaska at Fairbanks, have surprisingly small vectors, perhaps as a result of its remote location and unique circumstances.

This same convex-curve repeats across many countries, though with variations that can reveal characteristics about the scientific systems of each country (Fig.~\ref{fig:length}b). 
For example, Chinese organizations have a longer average embedding vector size ($\bar{l} = 4.81$) than organizations in the United States ($\bar{l} = 0.421$), reflecting the tendency for organizations in the United States to appear more frequently in more different contexts than Chinese organizations. 
Smaller but scientifically advanced countries such as Sweden, and other large populous countries like Brazil exhibit curves similar to the United States (Fig.~\ref{fig:length}b, inset), a pattern repeated across most of the 20 countries with the most researchers (Fig.~\ref{fig:concave20}). 
Other nations exhibit different curves, such as Egypt which has an "S"-shaped curve---their largest organizations do not have smaller vectors, as in many other countries. 
Other nations such as Poland and Turkey demonstrate a linear relationship between an organization's number of researchers and the vector size, indicating that their largest organizations belong to very specific contexts (Fig.~\ref{fig:concave20}).
By encoding whether an organization will appear in specific or many different contexts, organization embeddings offer a way to characterize universal and particularistic patterns of mobility across countries.

>>>>

Just as the embedding space makes it possible to ``zoom in'' on subsets of organizations, it is also possible to ``zoom out'' by aggregating organizational vectors. 
For example, we can examine the high-level structure by estimating the aggregate embeddings of countries. 
Namely, we can define the representative vector of each country as the average of their organizational vectors.
We then perform hierarchical clustering of nations that have at least 25 organizations represented in the embedding space, excluding the United States which is a dominant hub well connected to most other countries (see Fig. \ref{fig:clustering}a). 
The six identified clusters roughly correspond to countries in East-Asia (orange), Scandinavia (dark purple), the British Commonwealth (light purple), and Central and Eastern Europe (light blue), and then two remaining diverse clusters contain a mix of European, Latin American, and Mediterranean countries.
These clusters demonstrate the importance of not only geography but also linguistic ties between countries.

@@

Prestige hierarchy is known to underpin the dynamics of scientific mobility~\autocite{deville2014career, clauset2015hierarchy}, but can the hierarchy be encoded in the embedding space, to which no explicit information on prestige is given?
We test this by exploiting the geometric properties of the embedding space with SemAxis~\autocite{an2018semaxis}, an approach originally designed to characterize the semantics of words in word embeddings.
Here, we use SemAxis to operationalize the abstract notion of academic prestige, defining each pole of the axis using high- and low-ranked universities. 
We define prestige using the Times Ranking of World Universities (and using research impact from the Leiden Rankings~\autocite{waltman2012leidenrankings} in Supplementary Materials), including all ranked universities in the United States;
the high-rank pole is defined as the mean vector of the top five U.S. universities according to the rankings, whereas the low-ranked pole is defined using the five bottom-ranked and geographically matched (by U.S. census region) universities.
A geographic axis is also added to aid the visualization and reinforce the importance of geography, defined as the mean vector of all universities in California and then Massachusetts.

@@

The embedding space captures the prestige hierarchy and can be used to reconstruct academic rankings. 
We derive a ranking for universities using the SemAxis prestige axis in Fig.~\ref{fig:semaxis}a and compare them to traditional university rankings.
The embedding based ranking is defined from the position of organizations along the prestige axis, with the top-ranked being the furthest towards the high-ranked end of the axis.
We find that the embedding-based ranking is strongly correlated with the Times ranking (Spearman's $\rho = 0.73$, Fig.~\ref{fig:semaxis}e).
The strongest outliers that were ranked more highly in the Times Ranking than the embedding based ranking tend to be large state universities such as Arizona State University and  University of Florida;
Those ranked higher in the embedding based ranking tend to be relatively-smaller universities such as the University of San Francisco and University of Maryland Baltimore County that are close to other prestigious universities.

The correlation between the embedding based ranking and the Times ranking is robust with the number of organizations used to define the axes (Fig. SI~\ref{fig:supp:semaxis_compare}), such that even using only the single top-ranked and bottom-ranked universities are significantly correlated with the Times Ranking (Spearman's $\rho = 0.46$, Fig.~\ref{fig:supp:semaxis_compare}a).
For reference, the embedding-based ranking achieves correlation with the Times Ranking comparable to that between the Times ranking and the Leiden Ranking~\autocite{waltman2012leidenrankings}, a well-known bibliometrically-informed university ranking (Spearman's $\rho = 0.87$, Fig.~\ref{fig:semaxis}f).
This suggests that the organization embedding captures the latent prestige hierarchy inherent to academic mobility, even without providing any explicit information on prestige, with a high degree of resolution and reliability~\autocite{deville2014career, clauset2015hierarchy}.

What is it about the organization embedding that captures prestige so effectively?
We find that the size (L2 norm) of the organization embedding vectors provides insights into the characteristics of organizations. 
Up to a point, the size of U.S. organization's vectors tends to increase proportional to its number of researchers (both mobile and non-mobile) who published with that organization;
these organizations are primarily teaching-focused institutions, agencies, and hospitals that either are not ranked or have a low ranking. 
However, at around 1,000 researchers, the size of the vector decreases as the number of researchers increases;
organizations here are primarily research-intensive and prestigious universities with higher rankings, research outputs, R\&D funding, and doctoral students (Fig.~\ref{fig:supp:length_vs_metaInfo}).
A similar pattern has been observed in applications of neural embedding to natural language, wherein the size of word vectors was found to represent the word ``significance''~\autocite{schakel2015measuring}.
If a word appears frequently in a specific context, the norm of the word vector is likely to be large, because a large norm is associated with how predictably the word occurs given a context.
However, if the word in question is universal, appearing frequently in many different contexts, it cannot have a large norm as it lacks a strong association with a particular context. 
Likewise, an organization such as Harvard with a small norm appears in many contexts, alongside many different organizations in affiliation trajectories. 
The concavity can overlap into a single line, implying the size of the vector is related to the connectedness of the organization given its size (Fig.~\ref{fig:length}a, inset, $R^{2} = 0.17$).
Large, prestigious, and well-funded research universities such as Princeton and Harvard have smaller vector sizes because they appear in many different contexts compared to more teaching-focused organizations such as NY Medical College, and the University of Michigan at Flint. 
% Some universities, such as the University of Alaska at Fairbanks, have surprisingly small vectors, perhaps as a result of its remote location and unique circumstances. (I think it hurts the context)

This same convex-curve repeats across many countries, though with variations that can reveal characteristics about the scientific systems of each country (Fig.~\ref{fig:length}b). 
For example, Chinese organizations have a longer average embedding vector size ($\bar{l} = 4.81$) than organizations in the United States ($\bar{l} = 4.21$), reflecting the tendency for organizations in the United States to appear more frequently in more different contexts than Chinese organizations. 
Smaller but scientifically advanced countries such as Sweden, and other large populous countries like Brazil exhibit curves similar to the United States (Fig.~\ref{fig:length}b, inset), a pattern repeated across most of the 20 countries with the most researchers (Fig.~\ref{fig:concave20}). 
Other nations exhibit different curves, such as Egypt which has an ``S"-shaped curve---their largest organizations do not have smaller vectors, as in many other countries. 
Other nations such as Poland and Turkey demonstrate a linear relationship between an organization's number of researchers and the vector size, indicating that their largest organizations belong to specific contexts (Fig.~\ref{fig:concave20}).
%By encoding whether an organization will appear in specific or many different contexts, organization embeddings offer a way to characterize universal and particularistic patterns of mobility across countries.
